Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes

### 2024-02-01 Journal-manuscript/Journal-manuscript.tex
Revised manu draft 2- ready to submit back to reviewers

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal spatial availability: a singly-constrained
measure of accessibility considering multiple
modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
Antonio Páez\textsuperscript{\getAff{McMaster}}\\
\bigskip
\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
McMaster University, Hamilton, Canada\\
\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
\bigskip
* Corresponding author: soukhoa@mcmaster.ca (AS)\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Recent research has aimed to address the way opportunities are counted
in accessibility analysis. In conventional accessibility measures,
opportunities are often multiply counted, which leads to values of
accessibility that are difficult to interpret. Constraining the
calculations to match a known quantity ensures that the measurements sum
up to a predetermined quantity (i.e., the total number of
opportunities), and so each value can be meaningfully related to this
total. A recent effort is spatial availability, a singly-constrained
accessibility measure. In this paper we extend spatial availability for
use in the case of multiple modes or, more generally, heterogeneous
population segments with distinct travel behaviors. After deriving a
multimodal version of spatial availability, we proceed to illustrate its
features using a synthetic example. Next, we apply it to an empirical
example of low emission zones in Madrid, Spain. We conclude the paper
with suggestions for future research and its use in evaluating policy
interventions.

@@

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]} and is coming of age from the
perspective of planning research {[}see \emph{inter alia}, 4,5--8{]}.
Beginning with the work of Hansen {[}1{]}, accessibility measures have
been widely used to evaluate the efficiency of transportation systems
when combined with the distribution of populations and opportunities in
space {[}9{]}. In this way, accessibility is a holistic measure of
spatial systems' ease of reaching population-relevant destinations
{[}10,11{]}.

The most common form of accessibility measure is based the gravity model
{[}e.g. 1{]}: these measures sum ``weighted'' opportunities around a
focal point (i.e., a potential origin), based on how expensive it is to
reach them. Recent research in accessibility analysis has explored the
implications of methods by which opportunities are summed, especially in
the context of competitive accessibility {[}15{]}. In a typical
Hansen-type accessibility measure, the sums around origins are not
constrained. This means that the same opportunity can enter the sum for
different origins; Counting the same opportunity multiple times treats
the opportunity as if it was inexhaustible (or non-rival {[}16{]}).
However, this conceptualisation may not reflect reality as
opportunities, in general, are not inexhaustible. Some opportunities are
by definition exclusive: a typical example is employment, once a job is
taken up by someone in the population, the same job is no longer
available for any other person to take {[}14,17{]}. More generally, it
could be conceptualised that all opportunities are subject to some
amount of congestion or capacity-constraint: multiple people use a
specific opportunity at a given time and the more people who do, the
more congested the opportunity. Congestion can be seen on a continuum,
and debate on how congestion within accessibility should be considered
is ongoing {[}14,17,18{]}. Nonetheless, competition has been widely
considered for traditionally exclusive opportunities types such as
employment and healthcare opportunities {[}14,17,19--21,21--24{]} and
more recently educational facilities {[}18,25{]}, and less exclusive
amenities-types such as greenspace and recreation {[}25,26{]}, and
shopping {[}18,27{]}.

The consideration of congestion in accessibility measures was the
motivation for the development of approaches that consider competition
{[}21,23,24{]} and are popularly applied as floating catchment area
methods {[}e.g. 28{]}. While these approaches purport to account for
congestion, Paez et al. {[}12{]} demonstrates that they do not solve the
issue of multiple counting of opportunities in general, thus leading to
biases in the calculation of total demand and supply. They sometimes
inflate counts and other times deflate them. In response to this, recent
research has paid closer attention to the way opportunities are counted
in accessibility analysis. Paez et al. {[}12{]}, for example, focuses on
floating catchment area methods and introduces a normalization of the
impedance matrix to allocate the population and then the level of
service proportionally. More recently, Soukhov et al. {[}15{]}
introduced a singly-constrained measure of accessibility, called spatial
availability, that employs a similar but more sophisticated proportional
allocation mechanism. The work of these authors show that floating
catchment area methods can be seen as singly-constrained accessibility
measures and improve on existing approaches by guaranteeing that each
opportunity is counted only once. In other words, spatial availability
treats the opportunities as \emph{finite}. The proportional allocation
mechanism of spatial availability constrains the calculations to match a
known quantity, therefore ensuring that the measurements sum up to a
predetermined quantity (i.e., the total number of opportunities), and so
each value can be meaningfully related to this total.

A limitation of spatial availability as introduced by Soukhov et al.
{[}15{]} is that it was developed for the case of a homogeneous
population, for example for the case of a single mode of transportation.
However, the finite nature of opportunities makes the analysis of
heterogeneous populations very relevant. In the case of multiple modes
of transportation, people who travel by slower modes (e.g., active
modes) can usually reach fewer opportunities than people who travel by
faster modes and whose range is typically far wider (e.g., car). This
implies that slower travelers will often face increased competition for
local opportunities from travelers who can reach said opportunities from
farther afield.

This paper's primary motivation is to extend spatial availability for
the case of multimodal accessibility (i.e., travel by different modes).
It is worth noting though that the consideration for travel costs by
different modes is in fact just one case of heterogeneous populations.
The method itself can easily accommodate other forms of heterogeneity,
for example: variations in travel behavior between older and younger
adults {[}e.g., 29{]}, the propensity of older adults to use different
modes of transportation {[}e.g., 30{]}, the usually shorter trip lengths
of children compared to adults {[}e.g., 31{]}, or the more limited
travel ranges of single parents {[}e.g., 32{]}.

The rest of this paper is organized as follows. In Section 2 we provide
a brief review of multimodal accessibility. In the following Section 3
we demonstrate the derivation of the new spatial availability expression
for multiple modes. In Section 4 we illustrate relevant issues through a
synthetic example. This is followed in Section 5 by an empirical example
using home-to-work data from the city of Madrid, Spain after the
implementation of its Low Emission Zones (LEZ). Data for this example is
sourced from the city's 2018 travel survey. The empirical example
demonstrates the differences in multimodal spatial availability
highlighting differences within and outside the LEZ for travelers using
different modes, namely car, transit, cycling and walking. In Section 6,
we provide concluding remarks on the strengths of the use of spatial
availability as a multimodal accessibility measure, and discuss
potential future uses in policy planning scenarios as well as directions
for future research.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}33{]}; this includes cumulative
opportunity measures which are a special case of the gravity model
{[}5,10{]}. These gravity model based measures weight opportunities
depending on how easy it is to reach them. Given an origin (\(i\)) and a
destination (\(j\)), an impedance function \(f^{m}(c^m_{ij})\) converts
the cost of travel (e.g., time, money, generalized cost) into a score
that represents the propensity for potential interaction. These measures
originate from the work proposed by Hansen {[}1{]}, which can take the
following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\). In this form, \(m\) is a set of
modes that have mode-specific travel costs (\(c_{ij}^m\)) and/or travel
impedance functions (\(f^m(\cdot)\)).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al. {[}34{]} use Hansen-type accessibility to assess the potential
interaction with retail locations by three modes: walking, public
transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}34{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analysed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke {[}35{]}, researching
the region of Oslo, reports accessibility scores for car in the order of
tens of thousands of employment opportunities. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'', it is unclear what the unconstrained scores mean: is
having access to but 10,000 jobs by transit insufficient? After all,
10,000 employment opportunities are still plenty of opportunities. These
ratios can be found elsewhere in the literature {[}e.g., 36, and
9,29,32, and 8{]}, and they are useful as relative assessment of when
some members of the public are better or worse off than others, but they
are silent on how bad is ``worse off''.

Besides ratios of accessibility, another way seen in the literature to
improve interpretability of scores is to standardise them within a
{[}0-1{]} range. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardised in this way is not
uncommon in the literature {[}e.g., 37,38{]}.

Once we understand opportunities to be finite, it is possible for an
accessibility measure to take on a crisper meaning. As considered in the
long tradition of accessibility research, capacity of opportunities is
limited and thus is subject to competition by population
{[}14,15,17--25,39--42{]}. There are only so many school-seats, hospital
capacity, employment opportunities, etc., in a region and if one person
reaches an opportunity, it is taken: the supply of an opportunity and
the demand for that opportunity are two components of accessibility.
These are clear examples of opportunities that are unambiguously
competitive. But we would go as far as to argue that every type of
opportunity is subject to congestion or capacity constraint, even when
the opportunities are conventionally seen as non-competitive.

Amenities are a good example of this. For instance, standards for
providing green spaces are often stated in the form of \emph{exclusive
access}, in units of amenity per capita. A case in point is a the
Ile-de-France region, a jurisdiction that suggested in a 2013 planning
document that at the municipal level at least 10 \(m^2\) of public green
space should be supplied \emph{per inhabitant} {[}43{]}. Green spaces
are not evenly distributed, which means that who has access to them
hinges on where they are and how easy is to reach them. This formulation
of provision of amenities is not unusual. For example, Natural England,
an organization that recommends an Accessible Natural Greenspace
Standard such that the minimum supply of space is one ha of statutory
Local Nature Reserves per thousand population\footnote{see
  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
Similarly, the World Health Organization {[}cited in 44{]} recommends
that cities provide a minimum of 9 \(m^2\) of green area per inhabitant.
For our purposes, standards of this type translate into ``how much of
this resource is available to one individual that has not been claimed
by anyone else?''. Green spaces often have large capacities, but they
still have a capacity, and it is not the same for a person to have
access to 5 \(m^2\) of \emph{uncongested} green space as 15 \(m^2\).
This difference is in fact a matter of justice {[}43,45{]}. Constraining
accessibility is in this way a useful way to evaluate the congested
availability of any type of opportunity. As development of sound
standards is emphasized in the planning literature, in particular in
regards to fairness in transportation {[}see 46{]}, spatial availability
analysis is a useful way to develop and assess standards.

@@

As noted in the Introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by Luo et al. {[}28{]} who reformulated the work of Shen
{[}24{]} into two steps (although similar, and earlier, developments are
found in {[}21,23{]}). Shen-type accessibility is formulated as:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. Shen-type modal
accessibility (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
the travel impedance-weighted demand for opportunities. In this way, it
considers competition. That said, the measure remains unconstrained,
meaning both population \emph{and} opportunities are multiply counted
{[}see 12{]}. In other words, interpretation of the Shen-type
accessibility scores between modes is fraught, as it is for Hansen-type
measures.

To illustrate, {[}47{]} calculates \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m=\text{car}\). Their results
indicate that zones with low-income populations have lower \(a_i^m\)
than zones with higher-income populations. Further, they show that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many zones, arguing that this may further place those zones with
lower-income populations at a disadvantage. \(a_i\) and/or \(a_i^m\) are
used to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
opportunities were doubly counted (entering the sums of both modes),
this makes for uneasy interpretations of the differences in \(a_i^{m}\)
between modes. Questions that this approach leaves unaddressed include:
what is the impact of competition on the difference in \(a_i^m\) values?
How does the impact vary spatially? And what is the interpretation of
this difference?

Spatial availability improves on the discussed Hansen-type and Shen-type
accessibility approaches by constraining the sum of opportunities, that
is, by treating opportunities as finite. This is done by means of
proportional allocation factors that follow well established principles
of spatial interaction and the gravity model {[}see 48{]}. In Soukhov et
al. {[}15{]} these factors consider: the mass effect (e.g., the size of
populations at different origins); and the cost of travel from different
zones (e.g., some sub-populations face relatively higher or lower
costs). The following section introduces the multimodal form of spatial
availability.

@@

In brief, we define the spatial availability at an origin \(i\)
(\(V_{i}\)) as the proportion of all opportunities in the region that
are allocated to location \(i\) from all destinations \(j\). \(V_{i}\)
is a value of how many opportunities are available to each location
\(i\) out of all the opportunities in the region. The general
formulation of spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}) {[}see 15{]}:

@@

\begin{itemize}
\item
  \(F^t_{ij}\) is a balancing factor that depends on the size of the
  populations at different locations that demand opportunities \(O_j\),
  as well as the cost of movement in the system \(f(c_{ij})\).
\item
  \(V_i\) is the number of spatially available opportunities at \(i\);
  the sum of \(V_{i}\) is identical to the total number of opportunities
  in the region (i.e., \(\sum_j O_j = \sum_i V_i\)); in other words,
  opportunities are dealt with as finite resources.
\end{itemize}

Compared to Hansen-type accessibility
\(S_i = \sum_{j=1}^J O_jf(c_{ij})\), we see that spatial availability
is, like the Hansen-type measure, a weighted sum of the opportunities.
What makes spatial availability stand apart from other approaches is how
the weight used in the sum (balancing factor \(F^t_{ij}\)) implements a
proportional allocation mechanisms to ensure that the sum of \(V_i\) is
constrained to match the total number of opportunities in the region. As
such, spatial availability is singly-constrained and natively considers
competition. \(F^t_{ij}\) consists of two parts. The first part is a
population-based proportional allocation factor to model the mass effect
of the gravity model: \[
F^p_{i} = \frac{P_i}{\sum_i P_i}
\]

@@

Implementing \(F^{tm}_{ij}\), the following Equation
(\ref{eq:spatial-availability-multimodal}) gives the multimodal version
of spatial availability \(V_i^m\):

@@

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
shows a region with population at three population centers (\(A\),
\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
\(3\)). The population at each origin \(i\) is consists of two
sub-populations, one using a faster mode \(z\) and another using a
slower mode \(x\), to travel to employment centers. Population center
\(A\) is Suburban: it is closest to its own relatively large employment
center at \(1\), close to the Urban's equally large employment center
\(2\), and has a population that is smaller than the Urban \(B\) and
larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
population, followed by then \(A\), then \(C\). This synthetic example
is inspired by the single-mode example used in {[}24{]} and reconfigured
in {[}15{]}.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal spatial availability: a singly-constrained
measure of accessibility considering multiple
modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
Antonio Páez\textsuperscript{\getAff{McMaster}}\\
\bigskip
\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
McMaster University, Hamilton, Canada\\
\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
\bigskip
* Corresponding author: soukhoa@mcmaster.ca (AS)\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Place-based accessibility measures communicate the potential interaction
with opportunities at a zone that populations can access. Recent
research has explored the implications of how opportunities are counted
by different accessibility methods. In conventional measures,
opportunities are multiply counted if more than one zone offers access
to the same opportunity. This multi-count of opportunities leads to
values of accessibility that are difficult to interpret. A possible
solution to enhance the meaning-making of accessibility results is by
constraining the calculations to match a known quantity. This ensures
all zonal values sum up to a predetermined quantity (i.e., the total
number of opportunities). In this way, each value can be meaningfully
related to this total. A recent effort that implements this solution is
spatial availability, a singly-constrained accessibility measure. In
this paper, we extend spatial availability for use in the case of
multiple modes or more generally, heterogeneous population segments with
distinct travel behaviors. After deriving a multimodal version of
spatial availability, we proceed to illustrate its features using a
synthetic example. We then apply it to an empirical example of low
emission zones in Madrid, Spain. We conclude with suggestions for future
research and its use in evaluating policy interventions.

@@

Accessibility is a key concept in the analysis of land use and
transportation systems {[}1--3{]} and is coming of age from the
perspective of planning research {[}see \emph{inter alia} 4,5--8{]}.
Beginning with the work of Hansen (1959) {[}1{]}, accessibility measures
have been widely used to evaluate the efficiency of transportation
systems when combined with the distribution of populations and
opportunities in space {[}9{]}. In this way, accessibility is a holistic
measure of spatial systems' ease of reaching population-relevant
destinations {[}10,11{]}.

The most common form of accessibility measure is based on the gravity
model e.g.~the Hansen-type measure {[}1{]}: these measures sum
``weighted'' opportunities around a focal point (i.e., a potential
origin), based on how expensive it is to reach them. Recent research in
accessibility analysis has explored the implications of methods by which
opportunities are summed, especially in the context of competitive
accessibility {[}15{]}. In a typical Hansen-type accessibility measure,
the sums around origins are not constrained: the same opportunity can
enter the sum for different origins. Counting the same opportunity
multiple times treats it as it was inexhaustible (or non-rival
{[}16{]}). However, this conceptualisation may not reflect reality, as
opportunities are often more exclusive than not. Some opportunities are
by definition exclusive: a typical example is employment, once a job is
taken up by someone in the population, the same job is no longer
available for any other person to take {[}14,17{]}. More generally, it
could be conceptualised that all opportunities are somewhat exclusive so
they are subject to some amount of congestion or capacity-constraint:
multiple people use a specific opportunity at a given time and the more
people who do, the more congested the opportunity. Congestion can be
seen on a continuum, and debate on how congestion within accessibility
measures can be considered is ongoing {[}14,17,18{]}. Nonetheless,
competition has been widely considered for traditionally exclusive
opportunities types such as employment and healthcare opportunities
{[}14,17,19--21,21--24{]} and more recently for educational facilities
{[}18,25{]} as well as less exclusive amenities types such as greenspace
and recreation amenities {[}25,26{]}, and shopping destinations
{[}18,27{]}.

The consideration of congestion in accessibility measures was the
motivation for the development of approaches that consider competition
{[}21,23,24{]} and are popularly applied in the literature as floating
catchment area methods {[}28{]}. While these approaches purport to
account for congestion, Páez et al.~(2019) {[}12{]} demonstrates that
they do not solve the issue of multiple counting of opportunities in
general, thus leading to biases in the calculation of total demand
(population) and supply (of opportunities). They sometimes inflate
counts and other times deflate them. In this line, recent research has
paid closer attention to the way opportunities are counted in
accessibility analysis. Páez et al.~(2019) {[}12{]}, for example,
focuses on floating catchment area methods and introduces a
normalization of the impedance matrix to allocate the population and
then the level of service (i.e., the supply to demand) proportionally.
More recently, Soukhov et al.~(2023) {[}15{]} introduced a
singly-constrained measure of accessibility, called spatial
availability, that employs a similar but more sophisticated proportional
allocation mechanism. The work of these authors show that floating
catchment area methods can be seen as singly-constrained accessibility
measures and improve on existing approaches by guaranteeing that each
opportunity is counted only once. In other words, spatial availability
treats the opportunities as \emph{finite}. The proportional allocation
mechanism of spatial availability constrains the calculations to match a
known quantity, therefore ensuring that the measurements sum up to a
predetermined quantity (i.e., the total number of opportunities), and so
each value can be meaningfully related to this total.

A limitation of spatial availability as introduced by Soukhov et
al.~(2023) {[}15{]} is that it was developed for the case of a
homogeneous population, for example for the case of a single mode of
transportation. However, the finite nature of opportunities makes the
analysis of heterogeneous populations very relevant. In the case of
multiple modes of transportation, people who travel by slower modes
(e.g., active modes) can usually reach fewer opportunities than people
who travel by faster modes and whose range is typically far wider (e.g.,
car). This implies that slower travelers will often face increased
competition for local opportunities from travelers who can reach said
opportunities from farther afield.

This paper's primary motivation is to extend spatial availability for
the case of multimodal accessibility (i.e., travel by different modes).
It is worth noting though that the consideration for travel costs by
different modes is in fact just one case of heterogeneous populations.
The method itself can easily accommodate other forms of heterogeneity,
for example: variations in travel behavior between older and younger
adults (e.g., Páez et al.~(2010) {[}29{]}), the propensity of older
adults to use different modes of transportation (e.g., Moniruzzaman et
al.~(2013) {[}30{]}), the usually shorter trip lengths of children
compared to adults (e.g., Reyes et al.~(2014) {[}31{]}), or the more
limited travel ranges of single parents (e.g., Páez et al.~(2010)
{[}32{]}).

The rest of this paper is organized as follows. In Section 2, we provide
a brief review of multimodal accessibility. In the following Section 3,
we demonstrate the derivation of the spatial availability expression for
multiple modes. In Section 4, we illustrate relevant issues through a
synthetic example. This is followed in Section 5 by an empirical example
using home-to-work data from the city of Madrid, Spain after the
implementation of its Low Emission Zones (LEZ). Data for this example is
sourced from the city's 2018 travel survey. The empirical example
demonstrates the multimodal spatial availability landscape in 2018. It
highlights the differences within and outside the LEZ for travelers
using different modes, namely, car, transit, cycling and walking. In
Section 6, we provide concluding remarks on the strengths of the use of
spatial availability as a multimodal accessibility measure, and discuss
potential future uses in policy planning scenarios as well as directions
for future research.

@@

Place-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used
measures are based on the gravity model {[}33{]}; this includes
cumulative opportunity measures that are a special case of the gravity
model {[}5,10{]}. These gravity model based measures weight
opportunities depending on the ease of reaching them. Given an origin
\(i\) and a destination \(j\), an impedance function \(f^{m}(c^m_{ij})\)
converts the cost of travel (e.g., time, money, generalized cost) into a
score that represents the propensity for potential interaction. These
measures originate from the work proposed by Hansen (1959) {[}1{]},
which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\). In this form, \(m\) is a set of
modes that have mode-specific travel costs (\(c_{ij}^m\)) and/or travel
impedance functions \(f^m(\cdot)\).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al.~(2019) {[}34{]} use Hansen-type accessibility to assess the
potential interaction with retail locations by three modes: walking,
public transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of
retail locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}34{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\) i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analysed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke (2022) {[}35{]}
reports accessibility scores for car in the order of tens of thousands
of employment opportunities in the Oslo region. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'', it is unclear what the unconstrained scores mean: is
having access to 10,000 jobs by transit insufficient? After all, 10,000
employment opportunities are still plenty of opportunities. These ratios
can be found elsewhere in the literature e.g., Figs. 7, 8, and 9 in Páez
et al.~(2010) {[}36{]}, Fig. 5 in Páez et al.~(2010) {[}29{]}, Páez et
al.~(2013) Figs. 6, 7, and 8 in {[}32{]}. They are useful as relative
assessment of when some members of the public are better or worse off
than others, but they are silent on how bad is ``worse off''.

Besides ratios of accessibility, another way seen in the literature to
improve interpretability of scores is to standardise them within a
{[}0-1{]} range. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardised in this way is not
uncommon in the literature {[}37,38{]}.

If we understand opportunities to be finite and/or subject to some
levels of congestion, it is possible for an accessibility measure to
take on a crisper meaning. Accessibility research has a history of
considering opportunity competition, especially regarding school-seats,
hospital capacity, and employment opportunities
{[}14,15,17--25,39--42{]}. If one person reaches an opportunity - it is
taken: the supply of an opportunity and the demand for that opportunity
are the nodes in accessibility analysis. These types of opportunities
are unambiguously exclusive. But we would go as far as to argue that
every type of opportunity is subject to congestion or capacity
constraint, even when the opportunities are conventionally seen as
inexhaustible.

Amenities are a good example of this. For instance, standards for
providing green spaces are often stated in the form of \emph{exclusive
access}, in units of amenity per capita. For example, a 2013 planning
document for the Ile-de-France region suggested a public green space
municipal standard of 10 \(m^2\) \emph{per inhabitant} {[}43{]}. Green
spaces are not evenly distributed, meaning those who have access to them
depends on where they are and how easy they are to reach. This
formulation of amenity provision is not unusual. As another example,
Natural England recommends a national ``accessible natural greenspace
standard'' such that the minimum supply of space is 1 \(ha\) of
statutory local nature reserves per thousand population {[}44{]}.
Similarly, the World Health Organization {[}45{]} recommends that cities
provide a minimum of 9 \(m^2\) of green area per inhabitant. For our
purposes, standards of this type translate into ``how much of this
resource is available to one individual that has not been claimed by
anyone else?''. Green spaces often have large capacities, but they still
have a capacity and it is not the same for a person to have access to 5
\(m^2\) of \emph{uncongested} green space as 15 \(m^2\). This difference
is in fact a matter of justice {[}43,46{]}. Constraining accessibility
is a useful way to evaluate the congested availability of any type of
opportunity. As development of sound standards is emphasized in the
planning literature, in particular in regards to fairness in
transportation {[}47{]}, spatial availability analysis can be used to
develop and assess standards.

@@

As noted in the Introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by Luo et al.~(2003) {[}28{]} who reformulated the work of
Shen (1998) {[}24{]} into two steps (although similar, and earlier,
developments are found in {[}21,23{]}). Shen-type accessibility is
formulated as: \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\)
where \(D_j^m\) is the potential demand for opportunities equal to
travel impedance weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and
the remaining variables are repeated in the Hansen-type measure.
Shen-type modal accessibility (\(a_i^m\)) can be understood as a ratio
of the travel impedance-weighted supply of opportunities for \(m\)-mode
in \(i\) over the travel impedance-weighted demand for opportunities. In
this way, it considers competition. That said, the measure remains
unconstrained, meaning both population \emph{and} opportunities are
multiply counted {[}12{]}. In other words, interpretation of the
Shen-type accessibility scores between modes is fraught, as it is for
Hansen-type measures.

To illustrate, Tao et al.~(2020) {[}48{]} calculates \(a_i^m\) to jobs
for different income-group populations in Shenzhen, China for those
using transit and car. Their results indicate that zones with low-income
populations have lower \(a_i^m\) than zones with higher-income
populations. Further, they show that \(a_i^{\text{transit}}\) is lower
than \(a_i^{\text{car}}\) in many zones, arguing that this may further
place those zones with lower-income populations at a disadvantage.
\(a_i\) and/or \(a_i^m\) are used to compare relative spatial
differences in overall competitive accessibility and multimodal
competitive accessibility, but because opportunities were doubly counted
(entering the sums of both modes), this makes for uneasy interpretations
of the differences in \(a_i^{m}\) between modes. Questions that this
approach leaves unaddressed include: what is the impact of competition
on the difference in \(a_i^m\) values? How does the impact vary
spatially? And what is the interpretation of this difference?

Spatial availability improves on the discussed Hansen-type and Shen-type
accessibility approaches by constraining the sum of opportunities, that
is, by treating opportunities as finite. This is done by means of
proportional allocation factors that follow well established principles
of spatial interaction and the gravity model {[}49{]}. In Soukhov et
al.~(2023) {[}15{]} these factors consider: the mass effect (e.g., the
size of populations at different origins); and the cost of travel from
different zones (e.g., some sub-populations face relatively higher or
lower costs). The following section introduces the multimodal form of
spatial availability.

@@

In brief, we define the spatial availability \(V_{i}\) at an origin as
the proportion of all opportunities in the region that are allocated to
origin \(i\) from all destinations \(j\). The general formulation of
spatial availability is shown in Equation
(\ref{eq:spatial-availability-general}) {[}15{]}:

@@

\begin{itemize}
\item
  \(F^t_{ij}\) is a balancing factor that depends on the size of the
  populations at different locations that demand opportunities \(O_j\),
  as well as the cost of movement in the system \(f(c_{ij})\).
\item
  \(V_i\) is the number of spatially available opportunities at \(i\);
  the sum of \(V_{i}\) is identical to the total number of opportunities
  in the region (i.e., \(\sum_j O_j = \sum_i V_i\)).
\end{itemize}

Compared to Hansen-type accessibility
\(S_i = \sum_{j=1}^J O_jf(c_{ij})\), we see that spatial availability
is, like the Hansen-type measure, a weighted sum of the opportunities.
What makes spatial availability stand apart from other approaches is how
the weight used in the sum (balancing factor \(F^t_{ij}\)) implements a
proportional allocation mechanism to ensure that the sum of \(V_i\) is
constrained to match the total number of opportunities in the region. As
such, spatial availability is singly-constrained and natively considers
competition. \(F^t_{ij}\) consists of two parts. The first part is a
population-based proportional allocation factor to model the mass effect
of the gravity model: \[
F^p_{i} = \frac{P_i}{\sum_i P_i}
\]

@@

Implementing \(F^{tm}_{ij}\), Equation
(\ref{eq:spatial-availability-multimodal}) gives the multimodal version
of spatial availability \(V_i^m\):

@@

Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
shows a region with population at three population centers (\(A\),
\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
\(3\)). The population at each origin \(i\) is consists of two
sub-populations, one using a faster mode \(z\) and another using a
slower mode \(x\), to travel to employment centers. Population center
\(A\) is Suburban: it is closest to its own relatively large employment
center at \(1\), close to the Urban's equally large employment center
\(2\), and has a population that is smaller than the Urban \(B\) and
larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
population (40\%), followed by \(A\) (33\%) then \(C\) (30\%). This
synthetic example is inspired by the single-mode example used in Shen
(1998) {[}24{]} and reconfigured in Soukhov et al.~(2023) {[}15{]}.
